\MakeUppercase{The Structure of $k$-Planar Graphs}

### 2019-07-08 k-planar.tex
Changed width from dl(k^3+k^2+4k) to dl(k^3+3k)

\newcommand{\Bagsize}{\ensuremath{\frac{k^3}{6} + \frac{3k^2}{2} + \frac{13k}{3} + 4}}
% \newcommand{\bagsize}{\ensuremath{k^3/6 + 3k^2/2 + 13k/3 + 4}}
\newcommand{\treewidth}{\ensuremath{\binom{k+4}{3}}-1}

@@

\citet{dujmovic.joret.ea:planar} proved the following result, which with \cref{PartitionProduct}, implies \cref{PlanarBasic}.

\begin{thm}[\citep{dujmovic.joret.ea:planar}]
\label{PlanarPartition}
Every planar graph has:
\begin{compactitem}
\item an $H$-partition of layered width $1$ for some planar graph $H$ of treewidth at most $8$, and
\item an $H$-partition of layered width $3$ for some planar graph $H$ of treewidth at most $3$.
\end{compactitem}
\end{thm}

@@

\begin{thm}[\citep{dujmovic.joret.ea:planar}]
\label{SurfacePartition}
Every graph of Euler genus $g$ is a subgraph of:
\begin{compactenum}[(a)]
\item  $H \boxtimes P \boxtimes K_{\max\{2g,1\}}$ for some graph $H$ of treewidth at most $9$ and for some path $P$;
\item  $H \boxtimes P \boxtimes K_{\max\{2g,3\}}$ for some graph $H$ of treewidth at most $4$ and for some path $P$;
\item $(K_{2g} + H )  \boxtimes P$ for some graph $H$ of treewidth at most $8$  and some path $P$. 
\end{compactenum}
Equivalently, every graph of Euler genus $g$ has:
\begin{compactenum}[(a)]
\item an $H$-partition with layered width at most $\max\{2g,1\}$ such that $H$ has treewidth at most $9$;
\item an $H$-partition with layered width at most $\max\{2g,3\}$ such that $H$ has treewidth at most $4$; 
\item an $H$-partition with layered width at most $1$ such that $H$ has treewidth at most $2g+7$. 
\end{compactenum}
\textcolor{red}{eliminate some of these results? just keep (b)? }
\end{thm}

@@

\begin{thm}\label{modern-major-general}
  Let $G$ be a graph having a layered $H$-partition of layered width $\ell$ in which $H$ has treewidth at most $t$ and let $\PP$ be a $(k,d)$-shortcut system for $G$.  Then $G^\PP$ has a layered $J$-partition of layered width at most $d\ell(k^3+k^2+4k)$ for some graph $J$ with treewidth at most $\binom{k+t}{t}-1$.
\end{thm}

@@

\begin{proof}[Proof of \cref{modern-major-general}] 
  The proof is almost immediate, except that the layering $\mathcal{L}$ of $G$ may not be a valid layering of $G^{\mathcal{P}}$.  In particular, $G^{\mathcal{P}}$ may contain an edge $vw$ with $v\in V_i$ and $w\in V_{i+j}$ for any $j\in\{0,\ldots,k\}$.  To resolve this, we use a new layering $\mathcal{L}':=\langle V_0',\ldots,V_h'\rangle$ in which $V_i'=\bigcup_{j=ki}^{ki+k-1} V_i$.  This increases the layered width given by \clmref{general-width} from $d\ell(k^2+k+4)$ to $d\ell(k^3+k^2+4k)$.
\end{proof}

@@

Since $G_0$ is planar, the second part of \cref{PlanarPartition} implies that $G_0$ has an $H$-partition of layered width 3 for some planar graph $H$ of treewidth at most 3.  Applying \cref{modern-major-general} to $G_0$ and $\mathcal{P}$ immediately implies that $G$ (an arbitrary $k$-planar graph) has a layered $H$-partition of layered width $3(k^3+k^2+4k)$ for some graph $H$ of treewidth at most $\binom{k+4}{3}-1$.

@@

\begin{thm}
\label{PowerPartition}
Let $G$ be a graph of Euler genus $g$ and maximum degree $\Delta$. Let $k$ be a positive integer. 
Then $G^k$ has a layered $H$-partition of layered width at most $\max\{2g,3\} (2k^4+2k^3+8k^2) \Delta^{k} $ for some graph $J$ with treewidth at most $\binom{k+4}{4}-1$.
\end{thm}

@@

@@

We note that \cref{nearest-neighbour} is tight up to the leading constant:  Every $k$-nearest neighbour graph on $n$ vertices has at least $kn/2$ edges and at most $kn$ edges.  For $k\ge 8$, the Crossing Lemma~\citep{ajtai.chvatal.ea:crossing-free,leighton:complexity} implies that the total number of crossings is therefore $\Omega(k^3n)$ so that the average number of crossings per edge is at $\Omega(k^2)$.

\note{DW}{This actually works for $k\geq 7$, since the crossing lemma holds for graphs with at least $(3+\epsilon)n$ edges, for any fixed $\epsilon$. Of course, delete this comment.}

>>>>

%\newcommand{\Bagsize}{\ensuremath{\frac{k^3}{6} + \frac{3k^2}{2} + \frac{13k}{3} + 4}}
% \newcommand{\bagsize}{\ensuremath{k^3/6 + 3k^2/2 + 13k/3 + 4}}
\newcommand{\treewidth}{\ensuremath{\binom{k+4}{3}}-1}

@@

\citet{dujmovic.joret.ea:planar} proved the following result, which with \cref{MakeWidth1}, implies \cref{PlanarBasic}.

\begin{thm}[\citep{dujmovic.joret.ea:planar}]
\label{PlanarPartition}
Every planar graph has:
\begin{compactitem}
\item an $H$-partition of layered width $1$ for some planar graph $H$ of treewidth at most $8$, and
\item an $H$-partition of layered width $3$ for some planar graph $H$ of treewidth at most $3$.
\end{compactitem}
\end{thm}
Their proof is constructive and gives a simple quadratic-time algorithm for finding these partitions.

@@

\begin{thm}[\citep{dujmovic.joret.ea:planar}]
\label{SurfacePartition}
Every graph of Euler genus $g$ is a subgraph of:
\begin{compactenum}[(a)]
\item  $H \boxtimes P \boxtimes K_{\max\{2g,1\}}$ for some graph $H$ of treewidth at most $9$ and for some path $P$;
\item  $H \boxtimes P \boxtimes K_{\max\{2g,3\}}$ for some graph $H$ of treewidth at most $4$ and for some path $P$;
\item $(K_{2g} + H )  \boxtimes P$ for some graph $H$ of treewidth at most $8$  and some path $P$. 
\end{compactenum}
Equivalently, every graph of Euler genus $g$ has:
\begin{compactenum}[(a)]
\item an $H$-partition with layered width at most $\max\{2g,1\}$ such that $H$ has treewidth at most $9$;
\item an $H$-partition with layered width at most $\max\{2g,3\}$ such that $H$ has treewidth at most $4$; 
\item an $H$-partition with layered width at most $1$ such that $H$ has treewidth at most $2g+7$. 
\end{compactenum}
%PM: Keep 'em all: \textcolor{red}{eliminate some of these results? just keep (b)? }
\end{thm}

@@

\begin{thm}\label{modern-major-general}
  Let $G$ be a graph having a layered $H$-partition of layered width $\ell$ in which $H$ has treewidth at most $t$ and let $\PP$ be a $(k,d)$-shortcut system for $G$.  Then $G^\PP$ has a layered $J$-partition of layered width at most $d\ell(k^3+3k)$ for some graph $J$ with treewidth at most $\binom{k+t}{t}-1$.
\end{thm}

@@

\begin{proof}[Proof of \cref{modern-major-general}] 
  The proof is almost immediate, except that the layering $\mathcal{L}$ of $G$ may not be a valid layering of $G^{\mathcal{P}}$.  In particular, $G^{\mathcal{P}}$ may contain an edge $vw$ with $v\in V_i$ and $w\in V_{i+j}$ for any $j\in\{0,\ldots,k\}$.  To resolve this, we use a new layering $\mathcal{L}':=\langle V_0',\ldots,V_h'\rangle$ in which $V_i'=\bigcup_{j=ki}^{ki+k-1} V_i$.  This increases the layered width given by \clmref{general-width} from $d\ell(k^2+3)$ to $d\ell(k^3+3k)$.
\end{proof}

@@

Since $G_0$ is planar, the second part of \cref{PlanarPartition} implies that $G_0$ has an $H$-partition of layered width 3 for some planar graph $H$ of treewidth at most 3.  Applying \cref{modern-major-general} to $G_0$ and $\mathcal{P}$ immediately implies that $G$ (an arbitrary $k$-planar graph) has a layered $H$-partition of layered width $3(k^3+3k)$ for some graph $H$ of treewidth at most $\binom{k+4}{3}-1$.

@@

\begin{thm}
\label{PowerPartition}
Let $G$ be a graph of Euler genus $g$ and maximum degree $\Delta$. Let $k$ be a positive integer. 
Then $G^k$ has a layered $H$-partition of layered width at most $\max\{2g,3\} (2k^4+2k^3+8k^2) \Delta^{k} $ for some graph $J$ with treewidth at most $\binom{k+4}{4}-1$.
\end{thm}
\note{PM}{TODO: Deal with the change caused by the change to layered-width in \cref{modern-major-general}}.

@@

\note{PM}{TODO: Deal with the change caused by the change to layered-width in \cref{modern-major-general}}.

@@

We note that \cref{nearest-neighbour} is tight up to the leading constant:  Every $k$-nearest neighbour graph on $n$ vertices has at least $kn/2$ edges and at most $kn$ edges.  For $k\ge 7$, the Crossing Lemma~\citep{ajtai.chvatal.ea:crossing-free,leighton:complexity} implies that the total number of crossings is therefore $\Omega(k^3n)$ so that the average number of crossings per edge is at $\Omega(k^2)$.
